Data Analysis Methods for the Detection of the Epoch of Reionization

### 2017-12-01 PAPER.tex
small changes

By about one billion years after the Big Bang ($z \sim 6$), the very first stars and galaxies are thought to have ionized all the 
neutral hydrogen that dominated the baryonic matter content in the Universe. This transition period, during which the first 
luminous structures formed from gravitational collapse and began to emit intense radiation that ionized the cold neutral gas 
into a plasma, is known as the epoch of reionization (EoR). The EoR is a relatively unexplored era in our cosmic dawn. Its 
history encodes important information regarding the nature of the first galaxies and the processes of structure formation. 
Direct measurements of the EoR would unlock powerful information about the intergalactic medium, revealing connections 
between the matter distribution exhibited via cosmic microwave background (CMB) studies and the highly structured 
web of galaxies we observe today \citep[for a review, see][]{furlanetto_et_al2006}.

@@

The major challenge that faces all 21 cm experiments is isolating a small signal that is buried underneath foregrounds and 
instrumental systematics that are, when combined, four to five orders of magnitude brighter \citep[e.g.,][]{bernardi_et_al2009, pober_et_al2013, dillon_et_al2014, kohn_et_al2016}. A clean measurement therefore requires 
an intimate understanding of the instrument and a rigorous study of data analysis choices. With continual progress being made 
in the field and HERA on the horizon, it is becoming increasingly important to understand how the methods we choose interact 
with each other to affect power spectrum results. More specifically, it is imperative to develop techniques and tests that ensure 
the accuracy and reliability of a potential EoR detection. In this paper, we discuss three themes that are essential to investigate 
for a robust $21$ cm power spectrum analysis. We also highlight four power spectrum techniques and their trade-offs, potential 
pitfalls, and connections to the themes. We first approach the themes from a broad perspective, and then perform a detailed 
case study using data from the 64-element configuration of PAPER, motivating a revised PAPER-64 $21$ cm power spectrum 
from the lessons learned.

@@

In the next three sections, we investigate the effects of weighting matrices on signal loss by experimenting with different 
matrices $\textbf{R}$ and examine their impact on the resulting power spectra estimate $\hat{\textbf{p}}$. Our ultimate goal in 
experimenting with weighting is to suppress foregrounds and investigate EoR losses associated with it. We note that we 
purposely take a thorough and pedagogical approach to describing the toy model examples given in the next few sections, for 
the sake of completeness. The specifics of how signal loss appears in PAPER's analysis is later described in Section 
\ref{sec:Sigloss}.

@@

Since $\hat{\textbf{C}}$ is estimated from our data, its eigenspectrum differs from the eigenspectrum of our 
true $\textbf{C}$, and this difference has consequences on our result. An eigenspectrum ranks the eigenvalues of a matrix from 
highest to lowest and can be thought of as a spectrum of weights that are given to each spectral mode in the data. In other 
words, the eigenvalues encode the strength of different shapes in the dataset. The eigenspectrum of the identity matrix $
\textbf{I}$ is flat (all $1$'s) because it gives equal weighting to all modes. This is usually not the case for a covariance matrix, 
for which a sloped eigenspectrum means that modes are given different weights. The modes with the highest eigenvalues are 
down-weighted the most.

@@

In a $21$ cm power spectrum, detections could be the EoR signal, but they could also 
%(and unfortunately more likely) 
be attributed to other sources of bias. Connecting a detection to EoR as opposed to noise or foreground bias is a key challenge of 
future $21$ cm data analyses \citep[e.g.][]{petrovic_and_oh2011}. In this section we will discuss possible sources of bias in a measurement, as well as techniques 
that can help mitigate their effects. We will also present a series of tests in a pedagogical fashion which we suggest be used to 
help evaluate deep limits and/or detections.

@@

Because foreground spectra are smooth in frequency, they preferentially show up at low delay, or $k$ modes. For a particular 
baseline length, there is a maximum delay imposed on foregrounds, which corresponds to the light-crossing time between two 
antennas in a baseline. For longer baselines, this value increases, producing what is known as ``the 
wedge"
\citep{datta_et_al2010, parsons_et_al2012b, vedantham_et_al2012, pober_et_al2013, thyagarajan_et_al2013, liu_et_al2014a, liu_et_al2014b, patil_et_al2017}. 
The wedge describes a region in $k$-space contaminated by foregrounds, bounded by baseline length (which is proportional to $k_{\perp}$) and delay (which is 
proportional to $k_{\parallel}$). Properties of the wedge can be used to isolate and 
%remove 
avoid foregrounds, as done by P64, 
\citet{parsons_et_al2014}, and \citet{jacobs_et_al2015}.

However, this isolation is not perfect.  In deep measurements, power spectrum measurements at $k_{\parallel}$ values beyond 
the length of a baseline are often still contaminated at a low level. This leakage, especially at low $k$'s, can be attributed to 
convolution kernels associated with Fourier-transforming visibilities into delay-space. In other words, smooth-spectrum 
foregrounds appear as $\delta$-functions in delay-space, convolved by the Fourier transform of the source spectrum and the 
antenna response, both of which could smear out the foregrounds and cause leakage outside the wedge \citep[e.g.][]{ewall-wice_et_al2017}.

@@

As a brief review, PAPER is a dedicated $21$ cm experiment located in the Karoo Desert in South Africa. The PAPER-64 
configuration consists of 64 dual-polarization drift-scan elements that are arranged in a grid layout. For our case study, we 
focus solely on Stokes I estimated data \citep{moore_et_al2013} from PAPER's $30$ m East/West baselines (Figure 
\ref{fig:ant_layout}). For information about the backend system of PAPER-64, its observations, and data reduction pipeline, we 
refer the reader to \citet{parsons_et_al2010} and P64.

@@

This question is phrased visually in Figure \ref{fig:Pr_vs_Pin}. In the 2D space relating total $P_r$ to injected $P_{\rm in}$, very small injected levels begin at the level equivalent of no injection (i.e. data only). \cc{this next sentence does not make sense} As $P_{\rm in}$ increases, the signal loss transfer curve approaches the power level of the data it does not dominate the covariance but still contributes enough to suppressing itself. Above some level the injected signal is clearly detectable.

One difference in this method is that we are exclusively asking about signals detectable \emph{above} the level of the observed data. Thus it is explicitly limited to placing an upper limit, even if in some $k$-modes one has actual detections above the noise.  This might be seen as more conservative, however, a similar assumption has already been made when choosing to down-weight the data using itself. In such a down-weighting step, one has assumed that significant excess power is associated with a foreground residual. We note that a high SNR detection will require a different method, but we show this one to... \cc{fill in..}

@@

\section{Acknowledgements}
CC would like to acknowledge the UC Berkeley Chancellor's Fellowship, National Science Foundation Graduate Research 
Fellowship (Division of Graduate Education award 1106400), and thank Eric Switzer for helpful discussions. PAPER and HERA 
are supported by grants from the National Science Foundation (awards 1440343, and 1636646). ARP, DCJ, and JEA would 
also like to acknowledge NSF support (awards 1352519, 1401708, and 1455151, respectively). SAK is supported by the University of Pennsylvania School of Arts and Sciences Disseeration Completion Fellowship. We graciously thank SKA-
SA for site infrastructure and observing support.
\label{sec:Ack}

>>>>

By about one billion years after the Big Bang ($z \sim 6$), the very first stars and galaxies are thought to have ionized all the 
neutral hydrogen that dominated the baryonic matter content in the Universe. This transition period, during which the first 
luminous structures formed from gravitational collapse and began to emit intense radiation that ionized the cold neutral gas 
into a plasma, is known as the epoch of reionization (EoR). The EoR is a relatively unexplored era in our cosmic dawn. Its 
history encodes important information regarding the nature of the first galaxies and the processes of structure formation. 
Direct measurements of the EoR would unlock powerful information about the intergalactic medium, revealing connections 
between the matter distribution exhibited via cosmic microwave background (CMB) studies and the highly structured 
web of galaxies we observe today.

@@

The major challenge that faces all 21 cm experiments is isolating a small signal that is buried underneath foregrounds and 
instrumental systematics that are, when combined, four to five orders of magnitude brighter. A clean measurement therefore requires 
an intimate understanding of the instrument and a rigorous study of data analysis choices. With continual progress being made 
in the field and HERA on the horizon, it is becoming increasingly important to understand how the methods we choose interact 
with each other to affect power spectrum results. More specifically, it is imperative to develop techniques and tests that ensure 
the accuracy and reliability of a potential EoR detection. In this paper, we discuss three themes that are essential to investigate 
for a robust $21$ cm power spectrum analysis. We also highlight four power spectrum techniques and their trade-offs, potential 
pitfalls, and connections to the themes. We first approach the themes from a broad perspective, and then perform a detailed 
case study using data from the 64-element configuration of PAPER, motivating a revised PAPER-64 $21$ cm power spectrum 
from the lessons learned.

@@

In the next three sections, we investigate the effects of weighting matrices on signal loss by experimenting with different 
matrices $\textbf{R}$ and examine their impact on the resulting power spectra estimate $\hat{\textbf{p}}$. Our ultimate goal in 
experimenting with weighting is to suppress foregrounds and investigate EoR losses associated with it. We note that we 
purposely take a thorough and pedagogical approach to describing the toy model examples given in the next few sections, for 
the sake of completeness. The specifics of how signal loss crops up in PAPER's analysis is later described in Section 
\ref{sec:Sigloss}.

@@

It turns out that because we estimated $\hat{\textbf{C}}$ from our data, its eigenspectrum differs from the eigenspectrum of our 
true $\textbf{C}$, and this difference has consequences on our result. An eigenspectrum ranks the eigenvalues of a matrix from 
highest to lowest and can be thought of as a spectrum of weights that are given to each spectral mode in the data. In other 
words, the eigenvalues encode the strength of different shapes in the dataset. The eigenspectrum of the identity matrix $
\textbf{I}$ is flat (all $1$'s) because it gives equal weighting to all modes. This is usually not the case for a covariance matrix, 
for which a sloped eigenspectrum means that modes are given different weights. The modes with the highest eigenvalues are 
down-weighted the most.

@@

In a $21$ cm power spectrum, detections could be the EoR signal, but they could also (and unfortunately more likely) be 
attributed to other sources of bias. Connecting a detection to EoR as opposed to noise or foreground bias is a key challenge of 
future $21$ cm data analyses. In this section we will discuss possible sources of bias in a measurement, as well as techniques 
that can help mitigate their effects. We will also present a series of tests in a pedagogical fashion which we suggest be used to 
help evaluate deep limits and/or detections.

@@

Because foreground spectra are smooth in frequency, they preferentially show up at low delay, or $k$ modes. For a particular 
baseline length, there is a maximum delay imposed on foregrounds, which corresponds to the light-crossing time between two 
antennas in a baseline. For longer baselines, this value increases, producing what is known as ``the 
wedge" (\citealt{parsons_et_al2012b}; \citealt{liu_et_al2014a}; \citealt{liu_et_al2014b}; \citealt{vedantham_et_al2012}; 
\citealt{thyagarajan_et_al2013}; \citealt{pober_et_al2013}; \citealt{datta_et_al2010}). The wedge describes a region in $k$-
space contaminated by foregrounds, bounded by baseline length (which is proportional to $k_{\perp}$) and delay (which is 
proportional to $k_{\parallel}$). Properties of the wedge can be used to isolate and remove foregrounds, as done by P64, 
\citet{parsons_et_al2014}, and \citet{jacobs_et_al2015}.

However, this isolation is not perfect.  In deep measurements, power spectrum measurements at $k_{\parallel}$ values beyond 
the length of a baseline are often still contaminated at a low level. This leakage, especially at low $k$'s, can be attributed to 
convolution kernels associated with Fourier-transforming visibilities into delay-space. In other words, smooth-spectrum 
foregrounds appear as $\delta$-functions in delay-space, convolved by the Fourier transform of the source spectrum and the 
antenna response, both of which could smear out the foregrounds and cause leakage outside the wedge.

@@

As a brief review, PAPER is a dedicated $21$ cm experiment located in the Karoo Desert in South Africa. The PAPER-64 
configuration consists of 64 dual-polarization drift-scan elements that are arranged in a grid layout. For our case study, we 
focus solely on Stokes I estimated data (\citealt{moore_et_al2013}) from PAPER's $30$ m East/West baselines (Figure 
\ref{fig:ant_layout}). For information about the backend system of PAPER-64, its observations, and data reduction pipeline, we 
refer the reader to \citet{parsons_et_al2010} and P64.

@@

This question is phrased visually in Figure \ref{fig:Pr_vs_Pin}. In the 2D space relating total $P_r$ to injected $P_{\rm in}$, very small injected levels begin at the level equivalent of no injection (i.e. data only), very large levels are dominated by the injected signal, and a transition zone, where the injected signal amplitude is similar to the data level (and thus subject to cross correlations in both the covariance and in the power spectrum calculation).  It is in this last region that we search for the point at which an injected signal would be clearly detected.

One difference in this method is that we are exclusively asking about signals detectable \emph{above} the level of the observed data. Thus it is explicitly limited to placing an upper limit, even if in some $k$-modes one has actual detections above the noise.  This might be seen as more conservative, however, a similar assumption has already been made when choosing to down-weight the data using itself. In such a down-weighting step, one has assumed that significant excess power is associated with a foreground residual. We note that a high SNR detection will require a different method, but we show this one to... \cc{fill in..} \dcj{but we show this one to what?  Why not just leave it with a period}

@@

\section{Acknowledgements}
CC would like to acknowledge the UC Berkeley Chancellor's Fellowship, National Science Foundation Graduate Research 
Fellowship (Division of Graduate Education award 1106400), and thank Eric Switzer for helpful discussions. PAPER and HERA 
are supported by grants from the National Science Foundation (awards 1440343, and 1636646). ARP, DCJ, and JEA would 
also like to acknowledge NSF support (awards 1352519, 1401708, and 1455151, respectively). We also graciously thank SKA-
SA for site infrastructure and observing support.
\label{sec:Ack}
